---
title: "Diggins & Serti - JCI Insight - Figure 3"
output: html_notebook
author: "Kirsten Diggins"
---

This code was used to generate tSNE plots and heatmaps for Figure 3 of Diggins & Serti et al, "Exhausted-like CD8 T cell phenotypes linked to C-peptide preservation in alefacept-treated T1D subjects."

FCS files can be found https://www.itntrialshare.org/T1DAL.url. All additional code and annotation files can be found at: https://github.com/BenaroyaResearch/Diggins_Serti_Linsley_JCI_Insight_2020

```{r load_libraries, message=FALSE, warning=FALSE}
library(flowStats)
library(flowCore)
library(tidyr)
library(stringr)
library(ggplot2)
library(Rtsne)
library(grid)
library(viridis)
library(ggthemes)
library(dplyr)
library(reshape2)
library(ComplexHeatmap)

set.seed(123456)
```

## Load and process FCS files

```{r load_and_downsample_data, message=FALSE, warning=FALSE}
cd8_files <- list.files(path=path,pattern = "\\.fcs",recursive=T)
cd8_fcs <- lapply(paste0(path,cd8_files), read.FCS, transformation = F)

#### Process & transform data ####
process_data <- function(fcs){
  
  # Use marker names instead of dye names
  colnames(fcs) = make.names(toupper(pData(parameters(fcs))$desc))
  
  # Remove markers with multiple markers (i.e. all the NAs)
  fcs = fcs[,!(duplicated(colnames(fcs)) | duplicated(colnames(fcs), fromLast = T))]
  
  fcs = fcs[, order(colnames(fcs))]
}

cd8_fcs <- lapply(cd8_fcs, process_data)

## Transform the data
asinh_tfm_data <- function(fcs){
  # Arcsinh transform remaining columns
  tl <- transformList(colnames(fcs), arcsinhTransform(a=0, b=1/5), transformationId="asinh")
  fcs = transform(fcs, tl) 
}

cd8_arcsinh_trans <- lapply(cd8_fcs, asinh_tfm_data)

#### Extract and merge data (prep for pooled cytofkit analysis) ####
subjects <- lapply(cd8_arcsinh_trans,identifier) %>%
  unlist() %>%
  str_remove(.,".fcs") 
# subjects <- subjects[49:96]

mergedExpr = setNames(data.frame(matrix(ncol = ncol(exprs(cd8_arcsinh_trans[[1]]))+1, nrow = 0)), c(colnames(cd8_arcsinh_trans), "subj"))
for(i in 1:length(cd8_arcsinh_trans)){
  tmp.expr = as.data.frame(exprs(cd8_arcsinh_trans[[i]]))
  tmp.expr$subj = as.character(subjects[i])
  mergedExpr = rbind(mergedExpr, tmp.expr)
}

# Find the lowest number of CD8 events sampled
min_cd8_events <- min(table(mergedExpr$subj))

# Make a df with the same total number of total CD4 events for each subject (downsample to lowest CD8 level)
all_events_downsampled <- mergedExpr %>% group_by(subj) %>% do((function(x) {
  sample_idx <- sample(nrow(x), min_cd8_events)
  ds_df <- x[sample_idx,]
  return(ds_df)
})(.))

```

## tSNE analysis 

```{r run_tSNE}
all_data_subsample <- all_events_downsampled[sample(row.names(all_events_downsampled),100000,replace=F),]

cluster_markers <- c("CCR7", "CD38", "CXCR3", "CD27", "CD45RA", "CD45RO", "CD57","CD25", "TIGIT","EOMES", "T.BET", "CD95", "HELIOS", "KLRG1", "TIM3","PD1", "CD122", "X24B", "CD161","CD127","NKG2D")

tSNE_results <- Rtsne(all_data_subsample[,cluster_markers], dims=2, initial_dims = length(cluster_markers), check_duplicates = FALSE)

tSNE_coordinates = tSNE_results$Y
colnames(tSNE_coordinates) <- c("tSNE_1","tSNE_2")
data_with_tSNE <- cbind(as.data.frame(all_data_subsample),tSNE_coordinates)
```

## RPhenograph analysis

```{r run_Rphenograph}
data_to_cluster <- all_data_subsample[,c(cluster_markers)]

phenograph_results <- Rphenograph(data_to_cluster)

data_with_tSNE_and_clusters <- data.frame(data_with_tSNE,Cluster= phenograph_results$membership) 
```

## Generate Figures

Color points by marker expression:

```{r tSNE_plot_marker_expr}
data_with_tSNE_and_clusters <- tsne_week104 %>%
  # filter(Cluster %in% keep_clusts$Cluster)
  group_by(subj)

ggplot(data=data_with_tSNE_and_clusters,aes(x=tSNE_1,y=tSNE_2,col=as.factor(Cluster))) +
  geom_point(alpha=0.8) +
  # scale_color_manual(values=cols,name="Cluster") +
  # scale_color_viridis_c() +
  # scale_alpha(range = c(0.00, 0.5), guide = FALSE) +
  # stat_density2d(aes(alpha=..level..,fill=..level..),size=2,bins=10,geom="polygon") +
  # geom_density2d(color="black",bins=10) +
  # ylim(c(-11,11)) + xlim(c(-11,12))+
  coord_equal() +
  theme_classic()
```

Color only selected clusters and make the rest of the points gray:

```{r tSNE_plot_color_clusters}
### Color only selected clusters and make the rest of the points gray
cols = c("#FFFF99","#1F78B4","#FB9A99","#FDBF6F","#B2DF8A","#FF7F00","gray","#EF8A62","#CAB2D6","#E31A1C","#E9A3C9","black","#6A3D9A","#A6CEE3","#B15928", "#33A02C")

data_with_tSNE_and_clusters_select <- filter(data_with_tSNE_and_clusters,Cluster %in% c(12,20))
data_with_tSNE_and_clusters$new_ID <- rep(0,nrow(data_with_tSNE_and_clusters))

data_with_tSNE_and_clusters[!(data_with_tSNE_and_clusters$Cluster %in% c(12,20)),]$new_ID <- 0
data_with_tSNE_and_clusters[(data_with_tSNE_and_clusters$Cluster %in% c(12)),]$new_ID <- 12
data_with_tSNE_and_clusters[(data_with_tSNE_and_clusters$Cluster %in% c(20)),]$new_ID <- 20

ggplot(data=data_with_tSNE_and_clusters,aes(x=tSNE_1,y=tSNE_2)) +
  # geom_point(alpha=1,size=3)+
  geom_point(data=filter(data_with_tSNE_and_clusters,new_ID!=0),aes(x=tSNE_1,y=tSNE_2,col=as.factor(new_ID)),alpha=0.5,size=3)+
  scale_color_manual(values=c(cols[6],cols[13]),name="Cluster") +
  geom_density2d(color="gray27",linemitre=20) +
  coord_equal() +
  ylim(c(-11,11)) + xlim(c(-11,12))+
  theme_classic() 
```

Get cluster z-score heatmaps:

```{r marker_expr_heatmap}
plot_markers <- c("CCR7", "CD38", "CXCR3", "CD27", "CD45RA", "CD45RO", "CD57","CD25", "TIGIT","EOMES", "T.BET", "CD95", "HELIOS", "KLRG1", "TIM3","PD1", "CD122", "X24B", "CD161","CD127","NKG2D","GRZMB")
data_with_tSNE_and_clusters <- as.data.frame(data_with_tSNE_and_clusters)
cluster_medians <- stats::aggregate(. ~ data_with_tSNE_and_clusters$Cluster, data = data_with_tSNE_and_clusters[,c(plot_markers)], median) %>%
  dplyr::rename(Cluster = "data_with_tSNE_and_clusters$Cluster",`24B`= "X24B")

keep_clusts <- c(2,3,5,9,11,12,13,14,15,16,17,19,20,21,22,23)

cluster_medians <- cluster_medians %>%
  dplyr::filter(Cluster %in% keep_clusts)

col_scaled_median_RP <- scale(cluster_medians)
row.names(col_scaled_median_RP) <- cluster_medians$Cluster
colnames(col_scaled_median_RP)[colnames(col_scaled_median_RP)=="24B"] <- "2B4"

row.names(cluster_medians) <- cluster_medians$Cluster

cols = c("#FFFF99","#1F78B4","#FB9A99","#FDBF6F","#B2DF8A","#FF7F00","gray","#EF8A62","#CAB2D6","#E31A1C","#E9A3C9","black","#6A3D9A","#A6CEE3","#B15928", "#33A02C")
cluster_cols <- cols
names(cluster_cols) <- keep_clusts

heatmap_anno <- rowAnnotation(Cluster = row.names(col_scaled_median_RP), col=list(Cluster=c(col_scaled_median_RP)))

Heatmap(col_scaled_median_RP[,-1],row_names_side="right",name="Z-score",show_parent_dend_line=F, rect_gp = gpar(col = "white", lwd = 0.1)) #,right_annotation = heatmap_anno)
```



